---
title: "Chicago Budget Analysis"
author: "Michael Molloy"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    highlight: tango
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
    includes:
          after_body: "./assets/html/footer.html"
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r include=FALSE}
library(tidyverse)
library(magrittr)
```
#Chicago Emplpyee Data 
The City of Chicago publishes dozens of datasets every year through their data portal (https://data.cityofchicago.org/).  Since 2011, they have made public the salaries/ hourly rates for all City of Chicago employees.

# Loading Source Data
Source data was exported from the Chicago data portal into CSV format. The file was loaded into R as a dataframe. 
```{r}
# read data
Employee <- read_csv("./input/source/employees/Current_Employee_Names__Salaries__and_Position_Titles.csv")
```

# Data Wrangling
Before any analysis can be performed on the dataset we need to tidy up the data. 

```{r}
str(Employee)
```

## Mismatched Department Names
Over time, the city of Chicago has changed the exact text they use to describe departments.  While the numerical department code has stayed the same, it would be helpful if the department name text was consistent in our dataset. 

The graphic below shows the number of unique department names. 

```{r}
# are department names consistent?
Employee %>% 
  distinct(Department) 
```

To make it so the department names are consistent across years, we create a CSV file to map the original text and the text we want to change it to. Then a for loop is used to go through every record in the dataset to find and replace based on our mapping CSV file.

```{r}
# make department names consistent
  ## load the mapping table
  DepartmentNameMap <- read.csv("./input/mapping/Department Mapping Table.csv",stringsAsFactors = F)

  ## use the mapping table to replace the production data
    for (row in 1:NROW(DepartmentNameMap)) {
      from <- DepartmentNameMap[row,1]
      to <- DepartmentNameMap[row,2]
      Employee$Department <- str_replace(Employee$Department,from,to)
    }
```

## Mismatched Job Titles
For every position there is a human readable job title ("Title Description") and a unique code ("Title Code").  We can group the records by Title Code to see which codes have multiple human readable names.  The histogram below shows that while most title codes have only one human readable job title, some have two or three. These will need to be cleaned up.
```{r}
Employee %>% 
  distinct(Department) 
  
```

### Matching Approach
The first thing we do is capture all of the Job Codes that have multiple titles associated with them.

```{r}
JobTitleDupes <- budget %>% 
      distinct(`TITLE DESCRIPTION`,`TITLE CODE`) %>% 
      group_by(`TITLE CODE`) %>% 
      summarise(count = n()) %>% 
      filter(count >1) %>% # only codes with multiple job titles
      select(`TITLE CODE`)
```

Now we can take a look at the titles and see some easy patterns like one title being all upper case, others using abbreviations, and appending the title with the department name. 
```{r}
budget %>% 
      distinct(`TITLE DESCRIPTION`,`TITLE CODE`) %>% 
      group_by(`TITLE CODE`) %>% 
      filter(`TITLE CODE` %in% JobTitleDupes$`TITLE CODE`) %>% 
      arrange(`TITLE CODE`) %>% 
      head(15)
```

We can export the entire dataframe of Job Title duplicates to a CSV and delete the rows containing names we don't want, being sure to keep one row with the desired name.

Then, using a for loop, we can replace the Job Title for every record that matches the corresponding Title Code in our mapping file.

```{r}
 # make job titles consistent
  ## load the mapping table
 JobTitleMap <- read_csv("./input/source/mapping/Job Title Mapping Table.csv")


  ## use the mapping table to replace the production data
    for (row in 1:NROW(JobTitleMap)) {
      TitleDescription <- JobTitleMap[row,1]
      TitleCode <- JobTitleMap[row,2]
 budget[budget$`TITLE CODE` == as.character(TitleCode),"TITLE DESCRIPTION"] <- TitleDescription
    }
```


## Saving the Data
Once we've finished with our data wrangling we can save our tidy data to a single RDS file and begin our analysis.

```{r}
# write the output file
write_rds(budget,"./input/source/budget/Chicago Budget.Rds")
```